---
title: "Red Wine Quality: Exploratory Data Visualizations"
author: Brian Moore
output: rmarkdown::github_document
---

### Analysis scenario
- Use data visualizations to investigate which chemical properties 
correlate with higher quality red wine ratings.   

```{r echo=FALSE, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

### Read in the data
```{r}
options(scipen=999)
# warning: running this will install packages on your machine if they aren't present
required_packages <- c('tidyverse', 'dslabs', 'corrplot', 'gridExtra')
for(p in required_packages) {
  if(!require(p,character.only = TRUE)) 
        install.packages(p, repos = "http://cran.us.r-project.org")
  library(p,character.only = TRUE)
}
pct_formater_1 <- scales::label_percent(accuracy = 1)

url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-red.csv"

rw_df <- read_delim(url, delim=";")

# clean up column names (remove spaces in name)
rw_df <- rw_df %>% dplyr::rename_all(list(~make.names(.)))

# Uncomment lines below to learn more about the red wines dataset
# url_2 <- "https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality.names"
# read_lines(url_2)
```

### 10 example observations
```{r}
glimpse(head(rw_df, n=10))
```

### Summary stats on each of the data variables
```{r}
summary(rw_df)
```

### Quality counts
- The data dictionary indicates quality ratings could range from 0 to 10.
- Ratings only range from 3 to 8 for this dataset.
- Sparse values on the tails.
```{r}
rw_df %>% 
  count(quality) %>%
  mutate(percent_total = pct_formater_1(n/sum(n)))
```

### Create new groups for quality ratings
- Given the low and high end of the rating scale is less frequent we can build a new variable quality_2 so there's a higher frequency per quality level.
```{r}
rw_df <- rw_df %>%
      mutate(quality_2 = case_when(
            quality<=4 ~ "3 to 4",
            quality==5~ "5",
            quality==6 ~ "6",
            quality>6 ~ "7 to 8"),
      quality_2 = factor(quality_2, levels=c("3 to 4", "5", "6", "7 to 8")))

rw_df %>% 
  count(quality_2) %>%
  mutate(percent_total = pct_formater_1(n/sum(n)))
```

### Distribuion of each variable by quality rating

##### Observations on differences between groups:  
- Alcholol: looks like it tends to be higher for top rated wines.
- Citric Acid: lowest rating wines seem more likely to have lower citric acid.
- Sulphates: this metric seems to trend higher for top rated wines.
- Volatile Acidity: tends to be lower for top rated wines.

Histograms not preferred here due to the groups being uneven in size.  
In other words, showing counts via histograms would put the visual emphasis on the largest group size.
Density plots can be tricky for non-Data folks to grasp. See [Khan Academy: Density Curves](https://www.khanacademy.org/math/ap-statistics/density-curves-normal-distribution-ap/density-curves/v/density-curves) for a shareable resource to further intuition. 

```{r}
rw_df %>%
      select(-quality) %>%
      gather(key="metric", value="value", -quality_2) %>%
      ggplot(aes(x=value, fill=quality_2, color=quality_2)) +
      geom_density(alpha=0.7) +
      facet_wrap(metric ~ ., scale="free") +
      labs(title="Density Plots by Quality Rating and Metric",
           y="Density",
           x="Metric Value",
           fill="Quality Rating",
           color="Quality Rating") +
      theme(legend.position = "top")
```

### IQR plots
- Modified boxplots which don't include whiskers or outliers.
- Useful to compare the 25%, 50%, 75% quartiles of the data.
- e.g. chops the tails off the distribution.

##### Observations on differences between groups:  
- IQR plots reflect similar takeaways to the density plots.
- Slightly higher fixed acidity for top rated wines.
- Slightly lower PH for top rated wines.

```{r}
rw_df %>%
      select(-quality) %>%
      gather(key="metric", value="value", -quality_2) %>%
      group_by(quality_2, metric) %>%
      summarise(y0 = min(value),
                y25 = quantile(value, 0.25),
                y50 = median(value),
                y75 = quantile(value, 0.75),
                y100 = max(value), 
                .groups = 'drop') %>%
      ggplot(aes(x=quality_2, color=quality_2)) +
      geom_boxplot(
         # small hack here: set the min to q25 var and max to q75 var 
         # this prevents the whiskers from showing on the plot
         aes(ymin = y25, lower = y25, middle = y50, upper = y75, ymax = y75),
         stat = "identity") +
      facet_wrap(. ~ metric, scale="free") +
      labs(title="IQR Plots by Quality Rating and Metric",
           y="Metric Value",
           x="Rating",
           fill="Quality Rating",
           color="Quality Rating") +
      theme(legend.position = "none")
```

### Violin plots
- Violin plots offer another way to compare distributions across groups.
- These type of charts tend to be informative for a data audience and not obvious for non-Data folks.

```{r}
rw_df %>%
      select(-quality) %>%
      gather(key="metric", value="value", -quality_2) %>%
      ggplot(aes(x=value, y=quality_2, fill=quality_2, color=quality_2)) +
      geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
      facet_wrap(metric ~ ., scale="free") +
      labs(title="Violin Plots by Quality Rating and Metric",
           y="Quality Rating",
           x="Metric Value") +
      theme(legend.position = "none")
```

### Correlation plot: wine chemical metrics
- Investigate which metrics trend together using correlation matrix.
- Strongest correlation looks to exist between fixed acidity and pH metric.

```{r}
cor_mat <- cor(rw_df %>% select(-quality_2, -quality))

col <- colorRampPalette(c("#BB4444", "#EE9988", "#FFFFFF", "#77AADD", "#4477AA"))
corrplot(cor_mat, method="color", col=col(200),  
         type="upper", order="hclust", 
         addCoef.col = "black", 
         tl.col="black", tl.srt=45, #Text label color and rotation
         number.cex= 7/ncol(rw_df),
         diag=FALSE,
         title="Correlations between Wine Metrics",
         mar=c(0,0,1,0))
```

### Correlation plot: quality rating and wine metrics
- Investigate metric correlations with quality rating.

##### Observations on correlations with quality rating:  
- Alcohol has the strongest positive correlation with wine quality rating.
- Volatile acidity has the strongest negative correlation with wine quality rating.

```{r}
cor1 <- rw_df %>%
  select(-quality, -quality_2) %>%
  gather(key="metric", value="value") %>%
  group_by(metric) %>%
  summarise(correlation_with_quality = 
              cor(rw_df$quality, 
                  value),
            .groups="drop") %>%
  mutate(color_flag = ifelse(correlation_with_quality<0,"Below_0", "Above_0")) %>%
  ggplot(aes(y=reorder(metric, correlation_with_quality), 
             x=correlation_with_quality,
             fill=color_flag)) +
  geom_col(alpha=0.85) +
  scale_fill_manual(values=c("forestgreen", "salmon")) +
  theme(legend.position = "none") +
  labs(subtitle="Correlation with 
Wine Quality Rating",
       y="Metric",
       x="
Pearson Correlation Coefficient")

cor2 <- rw_df %>%
  select(-quality, -quality_2) %>%
  gather(key="metric", value="value") %>%
  group_by(metric) %>%
  summarise(correlation_with_quality = 
              cor(rw_df$quality, 
                  value),
            .groups="drop") %>%
  mutate(color_flag = ifelse(correlation_with_quality<0,"Below_0", "Above_0"),
         abs_cor = abs(correlation_with_quality)) %>%
  ggplot(aes(y=reorder(metric, abs_cor), 
             x=abs_cor,
             fill=color_flag)) +
  geom_col(alpha=0.85) +
  scale_fill_manual(values=c("forestgreen", "salmon")) +
  theme(legend.position = "none") +
  labs(subtitle ="Abs. Correlation with 
Wine Quality Rating",
       y="Metric",
       x="Absolute Value
Pearson Correlation Coefficient")

grid.arrange(cor1, cor2, ncol=2)
```

### Bivariate relationship: density and alcohol
- Top rated wines tend to have higher alcohol compared to other rating groups.
- Density is roughly similar between groups.
- Negative correlation between alcohol and density.

```{r}
rw_df %>% 
  group_by(quality_2) %>%
  mutate(median_density = median(density),
         median_alcohol = median(alcohol)) %>%
  ggplot(aes(x=density, y=alcohol)) +
  geom_point(aes(color=quality_2), alpha=0.4, show.legend = F) +
  geom_vline(aes(xintercept = median_density), linetype="dashed", color="grey40") + 
  geom_hline(aes(yintercept = median_alcohol), linetype="dashed", color="grey40") + 
  facet_wrap(. ~ quality_2, ncol=4) +
  theme(strip.text = element_text(size=16),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  labs(title="Rating Quality by Alcohol and Density",
       subtitle="Horizontal and vertical lines represent 
the rating group metric median.",
       y="Alcohol",
       x="Density")
```

### Bivariate relationship: citric acid and volatile acidity
- Seems higher quality wines clustered tend to have higher citric acid and lower volatile acidity.

```{r}
rw_df %>% 
  group_by(quality_2) %>%
  mutate(median_citric.acid = median(citric.acid),
         median_volatile.acidity = median(volatile.acidity)) %>%
  ggplot(aes(x=citric.acid, y=volatile.acidity)) +
  geom_point(aes(color=quality_2), alpha=0.4, show.legend = F) +
  geom_vline(aes(xintercept = median_citric.acid), linetype="dashed", color="grey40") + 
  geom_hline(aes(yintercept = median_volatile.acidity), linetype="dashed", color="grey40") + 
  facet_wrap(. ~ quality_2, ncol=4) +
  theme(strip.text = element_text(size=16),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  labs(title="Rating Quality by Citric Acid and Volatile Acidity",
       subtitle="Horizontal and vertical lines represent 
the rating group metric median.",
       y="Volatile Acidity",
       x="Citric Acid")
```

### Future ideas
- Is there a more robust wine library dataset that has features on soil, climate, wine maker, year, etc?
- Would grape region have a strong correlation with wine quality rating?
- Create a geo plot and overlay wine ratings on a map?
- Further explore how chemical properties interact?

### Data source citation
P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis.  
Modeling wine preferences by data mining from physicochemical properties.  
In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.  

Available at:  
[@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016  
[Pre-press (pdf)] http://www3.dsi.uminho.pt/pcortez/winequality09.pdf  
[bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib  

